---
title: "Bellabeat Data Analysis: Part 1 - Ask, Prepare and Process"
author: "Vladimir Rajevac"
date: "03/09/2021"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

# Background
Bellabeat, the manufacturer of wearable smart devices for women, believes analyzing how people are using their products may help unlock growth opportunities for the company.

Conducting the analysis will require looking into data how consumers use non-Bellabeat wearable smart devices and apply those insights to Bellabeat products.

### Objectives
* Identify trends in smart device usage and see how those trends might be applied to Bellabeat devices.
* Make recommendations, if any, that can be valuable in helping Bellabeat marketing strategy

### Data analysis process
This project is using the following data analysis steps: 

* Ask
* Prepare
* Process
* Analyze
* Share

# Ask
This section identifies the primary stakeholder and the business question/problem this data analysis will try to answer/offer a solution.

### Stakeholders
* **Urška Sršen**: Bellabeat’s cofounder and Chief Creative Officer
* **Sando Mur**: Mathematician and Bellabeat’s cofounder; key member of the Bellabeat executive team

### Business Task
Gain an understanding of how people use Fitbit smart devices and apply gained insights on one of the Bellabeat products to make recommendations that may help craft Bellabeat marketing strategy and grow the business.

# Prepare
For the analysis, we'll use [FitBit Fitness Tracker Data](https://www.kaggle.com/arashnic/fitbit) available at Kaggle by [Mobius](https://www.kaggle.com/arashnic). This data set is in the public domain. The data set contains personal tracker data collected from thirty FitBit users, including minute-level output for physical activity, heart rate, and sleep monitoring.

This section will explore FitBit data set, check how data is organized, and confirm its validity.

Install dependencies `tideverse` and `ggplot2`
```{r results="hide"}
install.packages('tidyverse')
install.packages('ggplot2')
library(tidyverse)
library(ggplot2)
```

### Explore Data Set

##### Data set: dailyActivity_merged.csv
```{r}
df_daily_activity_merged <- read_csv('raw-data/dailyActivity_merged.csv')
head(df_daily_activity_merged)
```

```{r}
str(df_daily_activity_merged)
```

```{r}
length(unique(df_daily_activity_merged$Id))
```
##### Data set: dailyCalories_merged.csv
This data set is part of `df_daily_activity_merged` data frame.
```{r}
df_daily_calories_merged <- read_csv('raw-data/dailyCalories_merged.csv')
head(df_daily_calories_merged)
```

```{r}
str(df_daily_calories_merged)
```

```{r}
length(unique(df_daily_calories_merged$Id))
```
##### Data set: dailyIntensities_merged.csv
This data set is part of `df_daily_activity_merged` data frame.
```{r}
df_daily_intensities_merged <- read_csv('raw-data/dailyIntensities_merged.csv')
head(df_daily_intensities_merged)
```
```{r}
str(df_daily_intensities_merged)
```

```{r}
length(unique(df_daily_intensities_merged$Id))
```
##### Data set: dailySteps_merged.csv
This data set is part of `df_daily_activity_merged` data frame. However, the number of daily steps might help find a relationship with other variables such as calories and METs.
```{r}
df_daily_steps_merged <- read_csv('raw-data/dailySteps_merged.csv')
head(df_daily_steps_merged)
```
```{r}
length(unique(df_daily_steps_merged$Id))
```
##### Data set: heartrate_seconds_merged.csv
We have heart rates from 14 of 33 unique FitBit users. The data is not part of the `df_daily_activity_merged` data frame.

```{r}
df_heartrate_seconds_merged <- read_csv('raw-data/heartrate_seconds_merged.csv')
head(df_heartrate_seconds_merged)
```
```{r}
str(df_heartrate_seconds_merged)
```

```{r}
length(unique(df_heartrate_seconds_merged$Id))
```

##### Data set: hourlyCalories_merged.csv
`hourlyCalories_merged.csv` might be used to understand at what time of the day users had the most intense training/activity.

```{r}
df_hourly_calories_merged <- read_csv('raw-data/hourlyCalories_merged.csv')
head(df_hourly_calories_merged)
```
```{r}
length(unique(df_hourly_calories_merged$Id))
```
```{r}
summary(df_hourly_calories_merged$Calories)
```

##### Data set: hourlyIntensities_merged.csv
It is not clear what `TotalIntensity` and `AverageIntensity` values represent. Since we shouldn't make the assumption, those variables shouldn't be used in the analysis. More information is needed to clarify this.


```{r}
df_hourly_intensities_merged <- read_csv('raw-data/hourlyIntensities_merged.csv')
head(df_hourly_intensities_merged)
```
```{r}
str(df_hourly_intensities_merged)
```
```{r}
unique(df_hourly_intensities_merged$TotalIntensity)
```
```{r}
unique(df_hourly_intensities_merged$AverageIntensity)
```
```{r}
max(df_hourly_intensities_merged$AverageIntensity)
min(df_hourly_intensities_merged$AverageIntensity)
```


##### Data set: hourlySteps_merged.csv
`hourlySteps_merged.csv` might be merged with `hourlyCalories_merged.csv` to check for correlation between the number of steps and calories per hour.


```{r}
df_hourly_steps_merged <- read_csv('raw-data/hourlySteps_merged.csv')
head(df_hourly_steps_merged)
```

```{r}
length(unique(df_hourly_steps_merged$Id))
summary(df_hourly_steps_merged$StepTotal)
```
##### Data set: sleepDay_merged.csv
This data set summs up daily sleaping patterns and it might be useful to make a connection between sleep and phisical activity.

```{r}
df_sleep_day_merged <- read_csv('raw-data/sleepDay_merged.csv')
head(df_sleep_day_merged)
```
```{r}
str(df_sleep_day_merged)
```

##### Data set: weightLogInfo_merged.csv
This data set might be used to find the correlation between sleep, daily activities, and weight loss.

```{r}
df_weight_merged <- read_csv('raw-data/weightLogInfo_merged.csv')
head(df_weight_merged)
```
```{r}
str(df_weight_merged)
```



### Data sets per minute observations
Side by `minuteMETsNarrow_merged.csv` data set, all other per minute observations may not suit the data analysis. Hourly observations should be sufficient to discover top-level patterns on how people are using FitBit devices.

##### Data set: minuteCaloriesNarrow_merged.csv

```{r}
df_minutes_calories_narrow_merged <- read_csv('raw-data/minuteCaloriesNarrow_merged.csv')
head(df_minutes_calories_narrow_merged)
```

```{r}
length(df_minutes_calories_narrow_merged$Id)
```

##### Data set: minuteCaloriesWide_merged.csv

```{r}
df_minutes_calories_wide_merged <- read_csv('raw-data/minuteCaloriesWide_merged.csv')
head(df_minutes_calories_wide_merged)
```

##### Data set: minuteIntensityNarrow_merged.csv
As with "hourlyIntensities_merged.csv" data set, it is not clear what Intensity means, and we should not make any assumptions, so this data set shouldn't be used.

```{r}
df_minutes_intensities_narrow_merged <- read_csv('raw-data/minuteIntensitiesNarrow_merged.csv')
head(df_minutes_intensities_narrow_merged)
```

```{r}
unique(df_minutes_intensities_narrow_merged$Intensity)
```

##### Data set: minuteIntensityWide_merged.csv
As with "hourlyIntensities_merged.csv" data set, it is not clear what Intensity means, and we should not make any assumptions, so this data set shouldn't be used.

```{r}
df_minutes_intensities_wide_merged <- read_csv('raw-data/minuteIntensitiesWide_merged.csv')
head(df_minutes_intensities_wide_merged)
```

##### Data set: minuteMETsNarrow_merged.csv
According to some information available at [Healthline](https://www.healthline.com/health/what-are-mets#definition) "One MET is the energy you spend sitting at rest — your resting or basal metabolic rate. So, an activity with a MET value of 4 means you’re exerting four times the energy than you would if you were sitting still.To put it in perspective, a brisk walk at 3 or 4 miles per hour has a value of 4 METs. Jumping rope, which is a more vigorous activity, has a MET value of 12.3."

While this data set might be used to track the most active times of the day, we can achieve the same by looking at hourly calories, intensities and steps.

```{r}
df_minutes_METs_narrow_merged <- read_csv('raw-data/minuteMETsNarrow_merged.csv')
head(df_minutes_METs_narrow_merged)
```
```{r}
unique(df_minutes_METs_narrow_merged$METs)
summary(df_minutes_METs_narrow_merged$METs)
```
##### Data set: minuteSleep_merged.csv
It is not clear what is the meaning of the `value` in the data set. More information is needed to use this data set.

```{r}
df_minutes_sleep_merged <- read_csv('raw-data/minuteSleep_merged.csv')
head(df_minutes_sleep_merged)
```
```{r}
unique(df_minutes_sleep_merged$value)
```
##### Data set: minuteStepsNarrow_merged.csv

```{r}
df_minutes_sleeps_narrow_merged <- read_csv('raw-data/minuteStepsNarrow_merged.csv')
head(df_minutes_sleeps_narrow_merged)
```

##### Data set: minuteStepsWide_merged.csv

```{r}
df_minutes_sleeps_wide_merged <- read_csv('raw-data/minuteStepsWide_merged.csv')
head(df_minutes_sleeps_wide_merged)
```
### Data set to use for the analysis
* dailyActivity_merged.csv
* hourlyCalories_merged.csv
* hourlyIntensities_merged.csv
* hourlySteps_merged.csv
* sleepDay_merged.csv
* weightLogInfo_merged.csv

###### Recommendations: ######
* Merge `sleepDay_merged.csv` with `dailyActivity_merged.csv`.
* Merge `hourlyCalories_merged.csv`, `hourlyIntensities_merged.csv`, and `hourlySteps_merged.csv`.
* Merge `weightLogInfo_merged.csv` with `dailyActivity_merged.csv`.

###### Data Validation and Limitations ######
It is not clear what was the criteria when collecting [FitBit Fitness Tracker Data](https://www.kaggle.com/arashnic/fitbit), how 33 observations are selected, and is the sample true representation of the population. One important thing to note is that Bellabeat customers are all female, and the FitBit data set doesn't make a distinction between male and female.

# Process

### Cleaning: dailyActivity_merged.csv

##### Check for data types #####
```{r}
df_daily_activity_merged <- read_csv('raw-data/dailyActivity_merged.csv')
str(df_daily_activity_merged )
```

Convert ActivityDate from character to date data type
```{r}
df_daily_activity_merged$ActivityDate <- as.Date(df_daily_activity_merged$ActivityDate, "%m/%d/%Y")
str(df_daily_activity_merged )
```

##### Remove duplicate rows from df_daily_activity_merged #####
Check for number of rows.
```{r}
nrow(df_daily_activity_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_daily_activity_merged <- distinct(df_daily_activity_merged)
nrow(df_daily_activity_merged)
```
Check for missing values
```{r}
colSums(is.na(df_daily_activity_merged))
```

Save cleaned data
```{r}
write.csv(df_daily_activity_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_daily_activity_clean.csv', row.names = TRUE)
```

### Cleaning: sleepDay_merged.csv

##### Check for data types #####
```{r}
df_sleep_day_merged <- read_csv('raw-data/sleepDay_merged.csv')
str(df_sleep_day_merged)
```
Convert SleepDay from character to date data type
```{r}
df_sleep_day_merged$SleepDay <- as.Date(df_sleep_day_merged$SleepDay, "%m/%d/%Y")
str(df_sleep_day_merged)
```

##### Remove duplicate rows from sleepDay_merged.csv #####
Check for number of rows.
```{r}
nrow(df_sleep_day_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_sleep_day_merged <- distinct(df_sleep_day_merged)
nrow(df_sleep_day_merged)
```
Check for missing values
```{r}
colSums(is.na(df_sleep_day_merged))
```

Save cleaned data
```{r}
write.csv(df_sleep_day_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_sleep_day_clean.csv', row.names = TRUE)
```

### Cleaning: weightLogInfo_merged.csv

##### Check for data types #####
```{r}
df_weight_merged <- read_csv('raw-data/weightLogInfo_merged.csv')
str(df_weight_merged)
```
Convert SleepDay from character to date data type
```{r}
df_weight_merged$Date <- as.Date(df_weight_merged$Date, "%m/%d/%Y")
str(df_weight_merged)
```

##### Remove duplicate rows from weightLogInfo_merged.csv #####
Check for number of rows.
```{r}
nrow(df_weight_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_weight_merged <- distinct(df_weight_merged)
nrow(df_weight_merged)
```
Check for missing values
```{r}
colSums(is.na(df_weight_merged))
```
Since there are 65 of 67 Fat variables with no value, we should drop this column.
```{r}
df_weight_merged <- select(df_weight_merged,-c(Fat))
colnames(df_weight_merged)
```

Save cleaned data
```{r}
write.csv(df_weight_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_weight_clean.csv', row.names = TRUE)
```

### Cleaning: hourlyCalories_merged.csv

##### Check for data types #####
```{r}
df_hourly_calories_merged <- read_csv('raw-data/hourlyCalories_merged.csv')
str(df_hourly_calories_merged)
```

Convert ActivityHour from character to date and time data type
```{r}

df_hourly_calories_merged$ActivityHour <- as.POSIXct(df_hourly_calories_merged$ActivityHour, 
                                                     format = "%m/%d/%Y %I:%M:%S %p")

str(df_hourly_calories_merged)
```
```{r}
head(df_hourly_calories_merged$ActivityHour, 24)
```


##### Remove duplicate rows from hourlyCalories_merged.csv #####
Check for number of rows.
```{r}
nrow(df_hourly_calories_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_hourly_calories_merged <- distinct(df_hourly_calories_merged)
nrow(df_hourly_calories_merged)
```
Check for missing values
```{r}
colSums(is.na(df_hourly_calories_merged))
```

Save cleaned data
```{r}
write.csv(df_hourly_calories_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_hourly_calories_clean.csv', row.names = TRUE)
```


### Cleaning: hourlyIntensities_merged.csv

##### Check for data types #####
```{r}
df_hourly_intensities_merged <- read_csv('raw-data/hourlyIntensities_merged.csv')
str(df_hourly_calories_merged)
```
```{r}
head(df_hourly_intensities_merged$ActivityHour, 24)
```


Convert ActivityHour from character to date and time data type
```{r}

df_hourly_intensities_merged$ActivityHour <- as.POSIXct(df_hourly_intensities_merged$ActivityHour, 
                                                     format = "%m/%d/%Y %I:%M:%S %p")

str(df_hourly_intensities_merged)
```
```{r}
head(df_hourly_intensities_merged$ActivityHour, 24)
```


##### Remove duplicate rows from hourlyIntensities_merged.csv #####
Check for number of rows.
```{r}
nrow(df_hourly_intensities_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_hourly_intensities_merged <- distinct(df_hourly_intensities_merged)
nrow(df_hourly_intensities_merged)
```
Check for missing values
```{r}
colSums(is.na(df_hourly_intensities_merged))
```

Save cleaned data
```{r}
write.csv(df_hourly_intensities_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_hourly_intensities_clean.csv', row.names = TRUE)
```


### Cleaning: hourlySteps_merged.csv

##### Check for data types #####
```{r}
df_hourly_steps_merged <- read_csv('raw-data/hourlySteps_merged.csv')
str(df_hourly_steps_merged)
```

Convert ActivityHour from character to date and time data type
```{r}

df_hourly_steps_merged$ActivityHour <- as.POSIXct(df_hourly_steps_merged$ActivityHour, 
                                                     format = "%m/%d/%Y %I:%M:%S %p")

str(df_hourly_steps_merged)
```
```{r}
head(df_hourly_steps_merged$ActivityHour, 24)
```


##### Remove duplicate rows from hourlySteps_merged.csv #####
Check for number of rows.
```{r}
nrow(df_hourly_steps_merged)
```
Remove duplicate rows, if any, and check for number of rows.
```{r}
df_hourly_steps_merged <- distinct(df_hourly_steps_merged)
nrow(df_hourly_steps_merged)
```
Check for missing values
```{r}
colSums(is.na(df_hourly_steps_merged))
```

Save cleaned data
```{r}
write.csv(df_hourly_steps_merged,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_hourly_steps_clean.csv', row.names = TRUE)
```

### Merge: `df_sleep_day_clean.csv` with `df_daily_activity_clean.csv`

```{r}
df_sleep_day_clean <- read_csv('clean-data/df_sleep_day_clean.csv')
df_daily_activity_clean <- read_csv('clean-data/df_daily_activity_clean.csv')
nrow(df_sleep_day_clean)
nrow(df_daily_activity_clean)
```

There are 410 observations in `df_sleep_day_clean.csv` and 960 in `df_daily_activity_clean.csv` 
```{r}
colnames(df_sleep_day_clean)
colnames(df_daily_activity_clean)
```
Rename SleepDay and ActiveDate columns to 'date'
```{r}
df_sleep_day_clean <- rename(df_sleep_day_clean, date = SleepDay)
df_daily_activity_clean <- rename(df_daily_activity_clean, date = ActivityDate)
colnames(df_sleep_day_clean)
colnames(df_daily_activity_clean)
```
Merge data frames
```{r}
df_sleap_activity_day <- merge(df_sleep_day_clean, df_daily_activity_clean, by=c('Id', 'date'))
nrow(sleap_activity_day)
```
Save data frame as .csv
```{r}
write.csv(df_sleap_activity_day,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_sleep_activity_day.csv', row.names = TRUE)
```

### Merge `df_hourly_calories_clean`, `df_hourly_intensities_clean.csv`, and `df_hourly_steps_clean.csv`.

```{r}
df_hourly_calories_clean <- read_csv('clean-data/df_hourly_calories_clean.csv')
df_hourly_intensities_clean <- read_csv('clean-data/df_hourly_intensities_clean.csv')
df_hourly_steps_clean <- read_csv('clean-data/df_hourly_steps_clean.csv')
nrow(df_hourly_calories_clean)
nrow(df_hourly_intensities_clean)
nrow(df_hourly_steps_clean)
```

```{r}
colnames(df_hourly_calories_clean)
colnames(df_hourly_intensities_clean)
colnames(df_hourly_steps_clean)
```

Merge data frames
```{r}
df_activity_hourly <- merge(df_hourly_calories_clean, df_hourly_intensities_clean, by=c('Id', 'ActivityHour'))
df_activity_hourly <- merge(df_activity_hourly, df_hourly_steps_clean, by=c('Id', 'ActivityHour'))
head(df_activity_hourly)
```
Save data frame as .csv
```{r}
write.csv(df_activity_hourly,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_activity_hourly.csv', row.names = TRUE)
```

### Merge: `df_weight_clean.csv` with `df_daily_activity_clean.csv`
```{r}
df_weight_clean <- read_csv('clean-data/df_weight_clean.csv')
df_daily_activity_clean <- read_csv('clean-data/df_daily_activity_clean.csv')
nrow(df_weight_clean)
nrow(df_daily_activity_clean)
```

There are 67 observations in `df_weight_clean.csv` and 960 in `df_daily_activity_clean.csv` 
```{r}
colnames(df_weight_clean)
colnames(df_daily_activity_clean)
```
Rename ActivityDate column in `df_daily_activity_clean.csv` to 'date'
```{r}
df_daily_activity_clean <- rename(df_daily_activity_clean, Date = ActivityDate)
colnames(df_daily_activity_clean)
```
Merge data frames
```{r}
df_weight_activity_day <- merge(df_weight_clean, df_daily_activity_clean, by=c('Id', 'Date'))
nrow(sleap_activity_day)
head(sleap_activity_day)
```
Save data frame as .csv
```{r}
write.csv(df_weight_activity_day,'~/Learning/Data\ Analytics/Google\ Coursera\ Data\ Analytics\ Course/7_Capstone/bellabeat-data-analysis/clean-data/df_weight_activity_day.csv', row.names = TRUE)
```
















